---
title: Depth distribution of OBIS records
date: "`r Sys.Date()`"
author: Pieter Provoost
output: (function(...) {
  rmdformats::robobook(toc_depth = 4, pandoc_args = c("+RTS", "-K2000m", "-RTS"), ...) })
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

## Dependencies

```{r message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(stars)
library(units)
library(ggplot2)
library(glue)
library(scales)
library(viridis)
```

## Calculate volume by depth zone: example

First download GEBCO GeoTiff data to `data/gebco_2021_sub_ice_topo_geotiff` and retile using `retile.sh``.

```{r message=FALSE, warning=FALSE}
gebco_files <- list.files("data/gebco_merged", pattern = "*_resampled_01.tif", full.names = TRUE)

dem <- stars::read_stars(gebco_files[1], proxy = FALSE)
names(dem) = "depth"
dem[["depth"]] <- as_units(dem[["depth"]], "m")

plot(dem, col = hcl.colors(11, "Spectral"), downsample = 0)
```

To calculate the water volume for a specific depth zone, restrict the DEM to values lower than or equal to the upper depth limit, then calculate the water column between the limits, and multiply with the surface area.

For example, for a depth zone between 2000 and 3000 meters:

```{r message=FALSE, warning=FALSE}
layer_top <- as_units(-2000, "m")
layer_bottom <- as_units(-3000, "m")
layer_height <- layer_top - layer_bottom

zone <- dem
zone[["depth"]][zone[["depth"]] > layer_top] <- as_units(NA, "m")

plot(zone, col = hcl.colors(11, "Spectral"), downsample = 0)
```

```{r message=FALSE, warning=FALSE}
zone$height <- layer_top - zone[["depth"]]
zone$height[zone$height > layer_height] <- layer_height

# stars bug?
# plot(st_downsample(zone, 20), attr = 2, col = hcl.colors(11, "Spectral"))

ggplot() +
  geom_stars(data = zone, aes(x, y, fill = height), downsample = 1) +
  scale_fill_distiller(palette = "Spectral", na.value = "#ffffff") +
  coord_sf()
```

Not sure how to calculate surface area with `stars`, move to `sf` for the remainder.

```{r message=FALSE, warning=FALSE}
zone_sf <- st_as_sf(zone, as_points = FALSE)
zone_sf$area <- st_area(zone_sf)
zone_sf$volume <- zone_sf$area * zone_sf$height
```

## Calculate volume for all depth zones

```{r message=FALSE, warning=FALSE}
depth_breaks <- c(seq(0, -200, by = -10), seq(-300, -1000, by = -100), seq(-2000, -11000, by = -1000))

results <- tibble()

for (i in 1:length(depth_breaks)) {
  layer_top <- as_units(depth_breaks[i], "m")
  layer_bottom <- as_units(depth_breaks[i + 1], "m")
  message(glue("{layer_top} -> {layer_bottom}"))
  layer_height <- layer_top - layer_bottom
  zone <- dem
  zone[["depth"]][zone[["depth"]] > layer_top] <- as_units(NA, "m")
  zone$height <- layer_top - zone[["depth"]]
  zone$height[zone$height > layer_height] <- layer_height
  zone_sf <- st_as_sf(zone, as_points = FALSE)
  remove(zone)
  zone_sf$area <- st_area(zone_sf)
  zone_sf$volume <- zone_sf$area * zone_sf$height
  results <- bind_rows(results, tibble(
    top = drop_units(layer_top),
    bottom = drop_units(layer_bottom),
    volume = sum(zone_sf$volume)
  ))
}
```

## Fetch OBIS data

```{r message=FALSE, warning=FALSE}
space <- S3FileSystem$create(
  anonymous = TRUE,
  scheme = "https",
  endpoint_override = "ams3.digitaloceanspaces.com"
)

#occ <- open_dataset(space$path("obis-datasets/exports/obis_20220221.parquet")) %>%
occ <- open_dataset("/Users/pieter/Downloads/obis_20221006.parquet") %>%
  select(minimumDepthInMeters, maximumDepthInMeters, year = date_year) %>%
  mutate(minimumDepthInMeters = as.numeric(minimumDepthInMeters), maximumDepthInMeters = as.numeric(maximumDepthInMeters)) %>%
  as_tibble()

occ$depth <- - rowMeans(occ[,c("minimumDepthInMeters", "maximumDepthInMeters")], na.rm = TRUE)

occ <- occ %>%
  filter(!is.nan(depth) & !is.na(year) & depth <= 0 & depth > -11000 & year >= 1800 & year < 2030) %>%
  select(depth, year)

depth_breaks <- c(seq(0, -200, by = -10), seq(-300, -1000, by = -100), seq(-2000, -11000, by = -1000))
year_breaks <- seq(1800, 2020, by = 10)

occ$depth_bin <- cut(occ$depth, breaks = depth_breaks, dig.lab = 10, right = TRUE)
occ$year_bin <- cut(occ$year, breaks = year_breaks, dig.lab = 10, right = FALSE)

stats <- occ %>%
  group_by(depth_bin, year_bin) %>%
  summarize(records = n())

stats$depth_start <- rev(depth_breaks)[stats$depth_bin]
stats$depth_end <- rev(depth_breaks)[as.numeric(stats$depth_bin) + 1]
stats$year_start <- year_breaks[stats$year_bin]
stats$year_end <- year_breaks[as.numeric(stats$year_bin) + 1]

reverselog_trans <- function() {
  trans <- function(x) -log10(x)
  inv <- function(x) 10^(-x)
  trans_new("revlog", trans, inv, log_breaks(base = 10), domain = c(1e-100, Inf))
}

ggplot() +
  geom_rect(data = stats, mapping = aes(xmin = year_start, xmax = year_end, ymin = -depth_start, ymax = -depth_end, fill = records)) +
  scale_fill_continuous(trans = "log10") +
  scale_y_continuous(trans = reverselog_trans())
```

```{r message=FALSE, warning=FALSE}
stats <- stats %>%
  left_join(results, by = c("depth_start" = "bottom", "depth_end" = "top")) %>%
  mutate(volume = volume / 1000000000) %>% # km3
  mutate(records = records / drop_units(volume))

ggplot() +
  geom_rect(data = stats, mapping = aes(xmin = year_start, xmax = year_end, ymin = -depth_start, ymax = -depth_end, fill = records)) +
  scale_fill_viridis(trans = "log10", option = "mako", direction = -1) +
  scale_y_continuous(trans = reverselog_trans()) +
  labs(fill = "Records per cubic km") +
  geom_hline(yintercept = 200, linetype = "dashed") +
  geom_hline(yintercept = 1000, linetype = "dashed") +
  geom_hline(yintercept = 4000, linetype = "dashed") +
  geom_hline(yintercept = 6000, linetype = "dashed") +
  geom_hline(yintercept = 11000, linetype = "dashed") +
  annotate(geom = "text", x = 1800, y = 200, label = "epipelagic", vjust = -1, hjust = 0) +
  annotate(geom = "text", x = 1800, y = 1000, label = "mesopelagic", vjust = -1, hjust = 0) +
  annotate(geom = "text", x = 1800, y = 4000, label = "bathypelagic", vjust = -1, hjust = 0) +
  annotate(geom = "text", x = 1800, y = 6000, label = "abyssopelagic", vjust = -1, hjust = 0) +
  annotate(geom = "text", x = 1800, y = 11000, label = "hadopelagic", vjust = -1, hjust = 0) +
  xlab("year") +
  ylab("depth")

ggsave("depth.png", width = 12, height = 7, scale = 1)
```







